---
title: "Example data preparation workflow"
author: "Robert Wildermuth"
date: "2025-9-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, loadLibraries}
library(rerddap)
library(tidyr)
library(lubridate)
library(here)
library(sf)
library(lubridate)
library(dplyr)
library(tidyr)
library(corrplot)

# Load helper functions to extract environmental predictors
# These rely on environmental netcdfs, which are stored in different places!
source("../R/getROMS.R") # ROMS ocean model variables
# source("./R/getMOM6_gridded.R") # MOM6 ocean model variables
source("../R/getCMEMS_l4chl.R") # CMEMS daily L4 chl
source("../R/getDistLand.R") # Distance to nearst land based on coast shp
source("../R/getBathym.R") # ETOPO bathymetry
```

# Prepare an example dataset for specied distribution modeling

Here we prepare a sample dataset to create a species distribution model for northern anchovy using the publicly available NOAA SWFSC California Current Ecosystem Survey (CCES) acoustic-trawl sampling observations.

First, we start by downloading and cleaning up the CCES data
```{r, CPStrawl}
# Download the latest SWFSC CPS trawl observations from ERDDAP
dataInfo <- info('FRDCPSTrawlLHHaulCatch', url = 'https://coastwatch.pfeg.noaa.gov/erddap/')
# Show the dataset columns (also see https://coastwatch.pfeg.noaa.gov/erddap/tabledap/FRDCPSTrawlLHHaulCatch.html)
cols <- info('FRDCPSTrawlLHHaulCatch')$variables
# Download the data
cps <- tabledap(dataInfo, fields = cols$variable_name)

# Process fields 
# NA values for subsamples mean zeroes
cps$subsample_count <- ifelse(is.na(cps$subsample_count), 0, cps$subsample_count)
cps$subsample_weight <- ifelse(is.na(cps$subsample_weight), 0, cps$subsample_weight)
# Use numbers and weights to show presence/absence
cps$pres <- ifelse(cps$subsample_count > 0 | cps$subsample_weight > 0, 1, 0)
# Get mean longitude and latitude
cps$lon <- rowMeans(cps[c("longitude", "stop_longitude")])
cps$lat <- rowMeans(cps[c("latitude", "stop_latitude")])
# plot(cps$lon, cps$lat) # Quick for any outlier locations
# Add a date column
cps$date <- as.Date(cps$time)
# Remove a few daytime samples
cps$hr <- hour(cps$time - hours(7)) # times originally in UTC
cps$dn <- ifelse(cps$hr < 6.1 | cps$hr > 17.9, "night", "day")
cps <- subset(cps, dn == "night")
# Convert to wide format
cpsMatrix <- pivot_wider(cps, names_from = scientific_name, values_from = pres, 
                         id_cols = c(cruise, haul, lon, lat, date, dn), values_fill = list(values = 0))
# Convert NA to 0 
cpsMatrix[is.na(cpsMatrix)] <- 0

# Add a few other useful columns
cpsMatrix$survey <- "cps"

# Subset to just anchovy
cpsAnch <- cpsMatrix[c("survey", "cruise", "haul", "lon", "lat", "date", "Engraulis mordax")]
colnames(cpsAnch)[ncol(cpsAnch)] <- "anchPA"

# Add a unique ID
cpsAnch$id <- seq(1:nrow(cpsAnch)) # used as 'obs' bellow
```

Next, we pull together physical covariate data from other public online databases.
```{r, getCovars}
# Load datasets: these currently live in project ./data
# coast <- sf::read_sf("./data/EPOCoast60_noGI.shp") # The coast shp
coast <- tigris::coastline() #!!RW: have to check if this will work for demonstration purposes
anch <- read.csv("./data/Hinchliffe_CSNA_timeseries_19652023.csv") # Anchovy SSB from Hincliffe et al. 2025

# To get a "persistence" forecast, we extract environmental variables from the year before the sampling date
# lag = 0 in the function extracts variables at the sampling date, lag = 1 extracts from the year before
# This is clunky but here I'm running the function twice to get lag = 0 and lag = 1, then joining the outputs
# Careful with lag > 0, could potentially give dates outside the temporal range of environmental datasets
#!!RW: generalize extractEnvVars() to specifiy which covariates to pull in
#      - OR just call the specific get*() fxns to make things more explicit
envExtractLag0 <- extractEnvVars(obs = obs, lag = 0)
envExtractLag1 <- extractEnvVars(obs = obs, lag = 1)
# Adjust colnames: clunky
#!!RW: Clean this up
colnames(envExtractLag1) <- paste0(colnames(envExtractLag1), "_lag1") # Adjust with lag number
# Figure out col index where environmental covariates start
startCol <- ncol(obs) + 1
envExtract <- cbind(envExtractLag0, envExtractLag1[, startCol: ncol(envExtractLag1)])
# Save the output
saveRDS(envExtract, "./data/combinedBioObs_envExtracted.rds")
# for now load from local file
envExtract <- readRDS("../data/combinedBioObs_envExtracted.rds")
```

We can investigate how these variables relate to the anchovy observations in the CCES dataset.
```{r}
# plot example time series
envExtract %>% ggplot(aes(x = year, y = sst)) +
  geom_point() +
  facet_wrap(~anchPA)

# plot correlations
corrMat <- cor(envExtract %>% select(lon, lat, anchPA, ild, sst, eke, chl, bathym, anchssb), use = "pairwise.complete.obs")
pTest <- cor.mtest(envExtract %>% select(lon, lat, anchPA, ild, sst, eke, chl, bathym, anchssb), alternative = "two.sided", method = "pearson")
corrplot(corrMat, p.mat = pTest$p, sig.level = 0.05, insig = "blank",
         order = 'hclust', hclust.method = "ward.D2", #"centroid", #"single", #
         tl.col = 'black', type = "lower",
         cl.ratio = 0.1, tl.srt = 45, tl.cex = 0.6, #mar = c(0.1, 0.1, 0.1, 0.1), 
         addrect = 6, rect.col = "green", diag = FALSE)

# Biplots of covariate space
envExtract %>% filter(anchPA == 1) %>%
  ggplot(aes(x = ild, y = sst)) +
  geom_point()
```

We will create a set of training and test subsets of the observational data for later use.
```{r}

```